---
title: "Forecasting Basics"
author: "*Carson Young*\n"
date: "February 2023"
output: 
  github_document:
    fig_width: 12
    fig_height: 9
classoption: a4paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include = FALSE}
library(tsibble)
library(tsibbledata)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(ggfortify)
library(feasts)
library(lubridate)
library(fpp3)
```


# Forecasting Introduction

Forecasting is about predicting the future as accurately as possible, given all of the information available, including historical data and knowledge of any future events that might impact the forecasts.

These notes are developed and adapted from Rob J Hyndman's *Forecasting: Principles and Practice*.


## `tsibble` objects 
A dataframe where one column is time and used as an index. Extends the standard `tibble` objects.

### `tsibble` basics
```{r}
# create tsibble object
(y <- tsibble(
  Year = 2015:2019,
  Observation = c(123, 39, 78, 52, 110),
  index = Year
))

```

```{r}
# create tibble object
(z <- tibble(
  Month = c("2019 Jan", "2019 Feb", "2019 Mar", "2019 Apr"),
  Observation = c(50,23,24,39),
))

```

```{r}
# mutate tibble to tsibble 
z %>%
  mutate(Month = yearmonth(Month)) %>%
  as_tsibble(index = Month)

```
Note the month column is converted from a text to a monthly time object.

## Olumpics Running Example
We can store multiple time series as a single `tsibble` object.
```{r}
data(olympic_running)
olympic_running
```
* [4Y] indicates the interval of these observations is every four years.
* There are 14 separate time series, see decompostition below. 
* No olympics in 1916 due to World War I, see missing value.

```{r}
olympic_running |> distinct(Sex)
olympic_running |> distinct(Length)
```
So seven different length categories

## Sales of pharmaceutical products in Australia Example
```{r}
data(PBS)
PBS
```
### `filter()` and `select()`
We can use the `filter()` function to extract the A10 scripts and the relavent column we want:
```{r}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost)

```
Note that the index variable `Month`, and the keys `Concession` and `Type`, would be returned even if they were not explicitly selected as they are required for a tsibble (to ensure each row contains a unique combination of keys and index).

### `summarise()`
The new variable TotalC is the sum of all Cost values for each month.
```{r}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(TotalC = sum(Cost))

```

### `mutate()`
Here we change the units from dollars to millions of dollars:
```{r}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(TotalC = sum(Cost)) |>
  mutate(Cost = TotalC/1e6) ->a10

```
Note right assignment at the end stpres the resulting tsibble. Useful operator after a long serires of piped functions. 

### Reading a csv file and convert to a tsibble

```{r}
prison <- readr::read_csv("https://OTexts.com/fpp3/extrafiles/prison_population.csv")
prison <- prison %>%
  mutate(Quarter = yearquarter(Date)) %>%
  select(-Date) %>%
  as_tsibble(key = c(State, Gender, Legal, Indigenous),
             index = Quarter)
prison

```
This tsibble contains 64 separate time series corresponding to the combinations of the 8 states, 2 genders, 2 legal statuses and 2 indigenous statuses. Each of these series is 48 observations in length, from 2005 Q1 to 2016 Q4.

For a tsibble to be valid, it requires a unique index for each combination of keys. 


\newpage

# Time series graphics 

## Time plots

```{r}
data(ansett)
melsyd_economy <- ansett |>
  filter(Airports == "MEL-SYD", Class == "Economy") |>
  mutate(Passengers = Passengers/1000)
autoplot(melsyd_economy, Passengers) +
  labs(title = "Ansett airlines economy class",
       subtitle = "Melbourne-Sydney",
       y = "Passengers ('000)")

```
`autoplot()` automatically produces an appropriate plot of whatever you pass to it in the first argument.

```{r}
autoplot(a10, Cost) +
  labs(y = "$ (millions)",
       title = "Australian antidiabetic drug sales")

```
Here, there is a clear and increasing trend. There is also a strong seasonal pattern that increases in size as the level of the series increases. The sudden drop at the start of each year is caused by a government subsidisation scheme that makes it cost-effective for patients to stockpile drugs at the end of the calendar year. Any forecasts of this series would need to capture the seasonal pattern, and the fact that the trend is changing slowly.

## Seasonal Plots
```{r}
a10 |>
  gg_season(Cost, labels = "both") +
  labs(y = "$ (millions)",
       title = "Seasonal plot: Antidiabetic drug sales")
```
Clear that there is a large jump in sales in January each year. Actually, these are probably sales in late December as customers stockpile before the end of the calendar year, but the sales are not registered with the government until a week or two later
```{r}
data(vic_elec)

# demand is a column in vic_elec dataframe
vic_elec |> gg_season(Demand, period = "day") +
  theme(legend.position = "none") +
  labs(y="MWh", title="Electricity demand: Victoria")

vic_elec |> gg_season(Demand, period = "week") +
  theme(legend.position = "none") +
  labs(y="MWh", title="Electricity demand: Victoria")

vic_elec |> gg_season(Demand, period = "year") +
  labs(y="MWh", title="Electricity demand: Victoria")
```
## Scatter plots

```{r}
vic_elec |>
  filter(year(Time) == 2014) |>
  autoplot(Demand) +
  labs(y = "GW",
       title = "Half-hourly electricity demand: Victoria")

vic_elec |>
  filter(year(Time) == 2014) |>
  autoplot(Temperature) +
  labs(
    y = "Degrees Celsius",
    title = "Half-hourly temperatures: Melbourne, Australia"
  )

vic_elec |>
  filter(year(Time) == 2014) |>
  ggplot(aes(x = Temperature, y = Demand)) +
  geom_point() +
  labs(x = "Temperature (degrees Celsius)",
       y = "Electricity demand (GW)")
```

## Scatterplot matricies

```{r}
data(tourism)
visitors <- tourism |>
  group_by(State) |>
  summarise(Trips = sum(Trips))
visitors |>
  ggplot(aes(x = Quarter, y = Trips)) +
  geom_line() +
  facet_grid(vars(State), scales = "free_y") +
  labs(title = "Australian domestic tourism",
       y= "Overnight trips ('000)")

library(GGally)
visitors |>
  pivot_wider(values_from=Trips, names_from=State) |>
  GGally::ggpairs(columns = 2:9)
```
\newpage

# Time series decomposition 

## Transformations and adjustments
### Population adjustments 
```{r}
global_economy |>
  filter(Country == "Australia") |>
  autoplot(GDP/Population) +
  labs(title= "GDP per capita", y = "$US")
```

### Inflation adjustments 
```{r}
print_retail <- aus_retail |>
  filter(Industry == "Newspaper and book retailing") |>
  group_by(Industry) |>
  index_by(Year = year(Month)) |>
  summarise(Turnover = sum(Turnover))
aus_economy <- global_economy |>
  filter(Code == "AUS")

print_retail |>
  left_join(aus_economy, by = "Year") |>
  mutate(Adjusted_turnover = Turnover / CPI * 100) |>
  pivot_longer(c(Turnover, Adjusted_turnover),
               values_to = "Turnover") |>
  mutate(name = factor(name,
         levels=c("Turnover","Adjusted_turnover"))) |>
  ggplot(aes(x = Year, y = Turnover)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y") +
  labs(title = "Turnover: Australian print media industry",
       y = "$AU")

```
Adjusted figures show Australia’s newspaper and book retailing industry has been in decline much longer than the original data suggests.

## Time series components

Additive decomposition

$$y_t = S_t + T_t + R_t$$

Multiplicative decomposition

$$y_t = S_t \times T_t \times R_t$$

where $y_t$ is the data,  $S_t$ is the seasonal component, $T_t$ is the 
trend-cycle component, and  
$R_t$ is the remainder component, all at period $t$.

## Example: Employment in the US retail sector

```{r}
data("us_employment")
us_retail_employment <- us_employment |>
  filter(year(Month) >= 1990, Title == "Retail Trade") |>
  select(-Series_ID)
autoplot(us_retail_employment, Employed) +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")
```
Try STL decomposition method. 

## STL decomposition 

```{r}
dcmp <- us_retail_employment |>
  model(stl = STL(Employed))
components(dcmp)
```


```{r}
components(dcmp) |>
  as_tsibble() |>
  autoplot(Employed, colour="gray") +
  geom_line(aes(y=trend), colour = "#D55E00") +
  labs(
    y = "Persons (thousands)",
    title = "Total employment in US retail"
  )
```

```{r}
components(dcmp) |> autoplot(colour = "#02597D")

```
STL has several advantages over classical decomposition, and the SEATS and X-11 methods described later:

- Unlike SEATS and X-11, STL will handle any type of seasonality, not only monthly and quarterly data.
- The seasonal component is allowed to change over time, and the rate of change can be controlled by the user.
- The smoothness of the trend-cycle can also be controlled by the user.
- It can be robust to outliers (i.e., the user can specify a robust decomposition), so that occasional unusual observations will not affect the estimates of the trend-cycle and seasonal components. They will, however, affect the remainder component.

### Seasonally adjusted data

If we remove seasonal component from the original data, the resulting values are "seasonally adjusted".

For the additive decomposition, this is simply $y_t-S_t$. i.e we only consider $T_t + R_t$

```{r}
components(dcmp) |>
  as_tsibble() |>
  autoplot(Employed, colour = "gray") +
  geom_line(aes(y=season_adjust), colour = "#0072B2") +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")

```
Useful technique when variation due to seasonality is not of primary interest. For example, monthly unemployment data are usually seasonally adjusted in order to highlight variation due to the underlying state of the economy rather than the seasonal variation.

## Moving average smoothing

A moving average of order $m$ can be written as


$$\hat{T_t} = \frac{1}{m}  \sum_{j = -k}^{k} y_{t+j} $$

where $m = 2k + 1$

The estimate of the trend-cycle at time $t$ is obtained by averaging values of the time series within $k$ periods of $t$. Observations that are nearby in time are also likely to be close in value. Therefore, the average eliminates some of the randomness in the data, leaving a smooth trend-cycle component, denoted by $m-\textbf{MA}$(moving avg of order $m$).

### Example Total Australian Exports 

```{r}
global_economy |>
  filter(Country == "Australia") |>
  autoplot(Exports) +
  labs(y = "% of GDP", title = "Total Australian exports")
```
Compute moving average of order 5 `slider` packages 
```{r}
library(slider)

aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  mutate(
    `5-MA` = slider::slide_dbl(Exports, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )

aus_exports |>
  autoplot(Exports) +
  geom_line(aes(y = `5-MA`), colour = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports with 5-MA estimate") +
  guides(colour = guide_legend(title = "series"))

```
```{r}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  mutate(
    `9-MA` = slider::slide_dbl(Exports, mean,
                .before = 4, .after = 4, .complete = TRUE)
  )

aus_exports |>
  autoplot(Exports) +
  geom_line(aes(y = `9-MA`), colour = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports with 9-MA estimate") +
  guides(colour = guide_legend(title = "series"))

```
The order of the moving average determines the smoothness of the trend-cycle estimate. In general, a larger order means a smoother curve.


#### Moving averages of moving averages

Consider $2 \times 4-\textbf{MA}$. i.e we apply $2-\textbf{MA}$ to the $4-\textbf{MA}$ 

$$
\begin{aligned}
\hat{T_t} &=\frac{1}{2} \left[\frac{1}{4}(y_{t-2}+y_{t-1}+y_t+y_{t+1}) + \frac{1}{4}(y_{t-1}+y_{t}+y_{t+1}+y_{t+2})\right]\\
&= \frac{1}{8}y_{t-2}+\frac{1}{4}y_{t-1}+\frac{1}{4}y_t+\frac{1}{4}y_{t+1}+\frac{1}{8}y_{t+2}
\end{aligned}
$$

```{r}
us_retail_employment_ma <- us_retail_employment |>
  mutate(
    `12-MA` = slider::slide_dbl(Employed, mean,
                .before = 5, .after = 6, .complete = TRUE),
    `2x12-MA` = slider::slide_dbl(`12-MA`, mean,
                .before = 1, .after = 0, .complete = TRUE)
  )
us_retail_employment_ma |>
  autoplot(Employed, colour = "gray") +
  geom_line(aes(y = `2x12-MA`), colour = "#D55E00") +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")

```
## Classical decomposition

```{r}
us_retail_employment |>
  model(
    classical_decomposition(Employed, type = "additive")
  ) |>
  components() |>
  autoplot() +
  labs(title = "Classical additive decomposition of total
                  US retail employment")

```

## X-11 method

X-11 uses a multiplicative decomposition, whereas the STL and classical decompositions are additive. 
```{r}
library(seasonal)
x11_dcmp <- us_retail_employment |>
  model(x11 = X_13ARIMA_SEATS(Employed ~ x11())) |>
  components()
autoplot(x11_dcmp) +
  labs(title =
    "Decomposition of total US retail employment using X-11.")
```
The X-11 trend-cycle has captured the sudden fall in the data due to the 2007–2008 global financial crisis better than either of the other two methods (where the effect of the crisis has leaked into the remainder component)
```{r}
x11_dcmp |>
  ggplot(aes(x = Month)) +
  geom_line(aes(y = Employed, colour = "Data")) +
  geom_line(aes(y = season_adjust,
                colour = "Seasonally Adjusted")) +
  geom_line(aes(y = trend, colour = "Trend")) +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail") +
  scale_colour_manual(
    values = c("gray", "#0072B2", "#D55E00"),
    breaks = c("Data", "Seasonally Adjusted", "Trend")
  )

```
The seasonally adjusted data is very similar to the trend-cycle component in this example, so it is hard to distinguish them on the plot.

## SEATS method

Seasonal Extraction in ARIMA Time Series (SEATS)
```{r}
seats_dcmp <- us_retail_employment |>
  model(seats = X_13ARIMA_SEATS(Employed ~ seats())) |>
  components()
autoplot(seats_dcmp) +
  labs(title =
    "Decomposition of total US retail employment using SEATS")

```


# Bibliography  
Hyndman, R. J., & Athanasopoulos, G. (2018). *Forecasting: principles and practice*. OTexts.